---
title: "Identifiability of beta and Q10"
author: "Robert Link"
date: "5/31/2019"
output:
  pdf_document: default
  html_document: default
---

## Introduction

Our purpose here is to see to what extent we can trade off the $\beta$ (CO$_2$
fertilization) parameter against the $Q_{10}$ (heterotrophic respiration enhancement)
parameter.  Both of these parameters affect the land-atmosphere carbon flux, but don't
(directly) affect anything else.  Therefore, if we can sterilize the effect of one of them
on land-atmosphere carbon flux by adjusting the other, the two are not independently
identifiable.  

## Structure of land-atmosphere flux
Write the carbon flux _from_ land _into_ atmosphere as
\begin{equation}
F = F_0 - N_0\beta \ln\left(\frac{C_a}{C_0}\right) + \alpha_d C_d Q_{10}^{T/10} + \alpha_s C_s Q_{10}^{\tilde{T}/10}.
\end{equation}
$F_0$ is the sum of the flux terms that don't involve either parameter.  $N_0$
is the NPP in the absence of CO$_2$ fertilization.  Hector assumes this to be
equal to the preindustrial value; it does not depend at all on the current model
state.  $C_a$ and $C_0$ are the current and base year atmospheric CO$_2$
concentrations.  The latter is approximately equal to the preindustrial value. 
$C_d$ and $C_s$ are the detritus and soil _carbon pools_, in PgC.  $T$ is
temperature, while $\tilde{T}$ is a trailing window-averaged temperature.  This
width of this window, in Hector's current configuration, is 200 years, meaning
that even in 2100 $\tilde{T}$ is the average temperature over the period
1900--2100.  Finally, $\alpha_d$ and $\alpha_s$ are constants, with value 0.025
and 0.02, respectively.  From time to time it will be convenient to write
$F_{d0} = \alpha_d C_d$ and $F_{s0} = \alpha_s C_s$.  These are the
detritus-atmosphere and soil-atmosphere fluxes in the absence of $Q_{10}$ enhancement.
Unlike $N_0$, they are not literally constant, since they depend on the size of the
carbon pools, but if the carbon pools are not changing significantly in size 
(an assumption we can check), then we can treat them as approximately constant.

From this expression we can get the partial derivatives with respect to $\beta$ and
$Q_{10}$.
\begin{align}
\frac{\partial F}{\partial \beta} &= -N_0 \ln\left(\frac{C_a}{C_0}\right). \\
\frac{\partial F}{\partial Q_{10}} &= \alpha_d C_d \frac{T}{10} Q_{10}^{T/10-1} + \alpha_s C_s \frac{\tilde{T}}{10}Q_{10}^{\tilde{T}/10-1}.
\end{align}

## Offset condition
Set
\begin{equation}
\frac{\partial F}{\partial Q_{10}} \Delta Q_{10} = - \frac{\partial F}{\partial \beta} \Delta \beta.
\end{equation}
Now solve for the ratio of $\Delta Q_{10}$ to $\Delta \beta$.
\begin{equation}
\frac{\Delta Q_{10}}{\Delta \beta} = \frac{10 N_0 \ln\left(\frac{C_a}{C_0}\right)}{\alpha_d C_d T Q_{10}^{T/10-1} + \alpha_s C_s \tilde{T} Q_{10}^{\tilde{T}/10 -1}}.
\end{equation}

We can rewrite this and simplify it by considering the limiting case where $\tilde{T} = 0$.
We get
\begin{equation}
\frac{\Delta Q_{10}}{\Delta \beta} = \frac{10N_0}{F_{d,0}} \frac{\ln\left(\frac{C_a}{C_0}\right)}{T}Q_{10}^{1-T/10}.
\end{equation}
However, if we are concerned that $\tilde{T} = 0$ is not a good approximation, we could instead
take $\tilde{T} = T$, as it would be if we set the window length to 0.  Then, we get
\begin{equation}
\frac{\Delta Q_{10}}{\Delta \beta} = \frac{10N_0}{F_{d,0} + F_{s,0}} \frac{\ln\left(\frac{C_a}{C_0}\right)}{T}Q_{10}^{1-T/10}.
\end{equation}
Notice that the dependence on $C_a$ and $T$ is exactly the same.  These represent the
two extremes of the windowing behavior, and so we expect them to bound the actual behavior.

## Can we offset beta and Q10?

The ratio of $\Delta Q_{10}$ to $\Delta \beta$ has some dependence on the atmosphere 
state variables, so we can't _exactly_ offset the two, but we need to ask ourselves how
close we can come.  The first term is (by an assumption that we still need to check) constant 
or nearly so, so all the variation comes from the second term

Here is a plot of the second term over time.

```{r ratio}
library('hector')
inifile <- system.file('input/hector_rcp85.ini', package='hector')
rfac <- function(t) 
{
    hcore <- newcore(inifile)
    run(hcore, max(t))
    c0 <- fetchvars(hcore, NA, PREINDUSTRIAL_CO2())$value
    ca <- fetchvars(hcore, t, ATMOSPHERIC_CO2())$value
    carbonfac <- log(ca/c0)
    
    temp <- fetchvars(hcore, t, GLOBAL_TEMP())$value
    q10 <- fetchvars(hcore, NA, Q10_RH())$value
    tempfac <- q10^(1-temp/10) / temp
    
    carbonfac*tempfac
}
curve(rfac, from=1900, to=2100)
```
There are some spikes in the historical period that are probably due to rapid
changes of temperature around volcanic events.  Outside of that, the values are
not changing a lot.

Here's what it looks like over the 21st century.
```{r ratio2}
curve(rfac, from=2000, to=2100)
```

So, if you took a compromise value of, say, 0.45, you might sometimes be making 
an error as high as 0.07/0.38 = 18% relative to the optimal value of $Q_{10}$
for that time step.  That's not nothing, but that's also as big as it gets.  For 
comparison, changing $Q_{10}$ by 18% from 2 to 2.36 results in a difference in
atmospheric CO$_2$ _accumulated over the entire run_ of `r signif(906.3063 - 888.0929, 2)`
ppm (906.3 as compared to 888.1).  If we assume that this error occurred entirely
over the last 10 years (which is surely not the case), we're looking at an error of 
`r signif((906.3063 - 888.0929)/10, 2)` ppm in each year, which I'm pretty sure
we couldn't see in our optimization.

## What next?
We still need to do our numerical tests to see if these estimates are borne out.
I'd be particularly interested in seeing how close our $\Delta Q_{10} / \Delta \beta$ 
estimate was.  However, assuming the story holds up, then what do we do next?

It looks to me like the only parameters that could disambiguate these two are the
individual components of the land flux, NPP and RH, since they don't directly enter
into any other calculations, and all of their indirect effects seem to happen through
the atmospheric concentration.  However, given how simplistic
Hector's calculations of NPP and RH are, I wonder to what extent we can really
expect to match them, even assuming they are available from the ESMs in the first
place.

If we can't get NPP or RH for the ESMs, or if we can't get the model to match
them individually, then we may have to just accept that we only really have one
free parameter here.
