---
title: "PCA First"
output: html_notebook
---


## Objective 

I am trying to look into the questions you posed in the "pca-explore.pdf" from last week. 

1. Does the story hold up when we extend to more scenarios or more variables?
2. Can we used this to build a credible argument that our mesa function methodology makes as much sense in the principal components coordinate system as it does in the natural coordinate system?

So far most of my efforts have been focused on question number 1 looking into the the following... 

* A. What happens when we increase the number of years going into the pca?  
* B. What happens when we increase the size of the Hector results ensemble?  
* C. What happens when we increase the number of scnearios going into the PCA? 
    + Do we need to scale? 
* D. What happens when we do the PCA on both temp and CO2? 

*** 

```{r, echo = FALSE, message = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(hector)
library(hectorcal)
library(foreach)

knitr::opts_chunk$set(echo = TRUE, fig.width = 10)

# Set the working direcotry for this script to the project location
BASE <- '..'
```


# Recreate the "explore-pca.pdf"

I wanted to start out at the same place that the explore-pca pdf left off at. I set my seed to the same one used in the pdf.

Emissions driven RCP 85 with oonly 250 Hector cases every 10 years from 1850 to 2100. 

**Expectation:** It should match the explore-pca.pdf 

```{r}
# Start by importing, checking and formatting the emissions driven run results. 
path <- list.files(file.path(BASE, 'data-raw', 'PCA_hector_ensemble'), 'emission_rcp85', full.names = TRUE)
rcp85_str <- readRDS(path)

# Foramt into as a wide tibble where one row represents on "case" of Hector.
tibble::as_tibble(rcp85_str$rslt) %>% 
    dplyr::filter(variable == 1) %>% 
    dplyr::mutate(index = paste0('rcp85_', variable, '_', year)) %>% 
    dplyr::select(run_id, index, value) %>% 
    tidyr::spread(index, value) %>%  
    na.omit %>% 
    dplyr::select(-run_id) -> 
    temp_rcp85_results

# We do not scale the results with this becasue we are jsut trying to recreate 
pca_temp_rcp85 <- prcomp(temp_rcp85_results[1:250, seq(from = 1, to = 241, by = 10)], 
                         center = TRUE, scale. = FALSE)
```

How much of the variability is explained by the different pcs? 

```{r}
fvar_temp_rcp85 <- calc_variance(pca_temp_rcp85)
head(fvar_temp_rcp85)

plot_fvar_temp_rcp85 <- plot_frac(fvar_temp_rcp85, title = 'Hector Emission Driven PCA RCP 8.5 Only') + 
    ggplot2::labs(subtitle = 'recreate RPL example')

plot_fvar_temp_rcp85

```

What do the first 3 pcs look like?  (I only use the first 250 results and I subset the time series so that it only includes results from every 10 years)
```{r}
plot_pcs(pca_temp_rcp85, 1:3, title = 'Hector Emission Driven Tgav PCA RCP 8.5 Only')  -> 
    plot_pc_rcp85;

plot_pc_rcp85 + 
    ggplot2::labs(subtitle = 'recreate RPL')
```

**Conclusion:** Awesome it looks like the exploratory one! 

<br>

## A. Increase the number of years

What happens when we add more years to the PCA? So we still only have 250 obs.

**Expectation:** I think that the variance should still look the same and that the PCs will be a little smoother. 

```{r}
pca_temp_rcp85_A <- prcomp(temp_rcp85_results[1:250, ], center = TRUE, scale. = FALSE)

fvar_temp_rcp85_A <- calc_variance(pca_temp_rcp85_A)

plot_frac(fvar_temp_rcp85_A, title = 'Hector Emission Driven PCA RCP 8.5 Only') + 
    labs(subtitle = 'All the Years (only showing the 1st 15 PCs)') + 
    coord_cartesian(xlim = c(0, 15))-> 
    plot_fvar_temp_rcp85_A; plot_fvar_temp_rcp85_A
```


How does the fraction of the variance captured by the PCs change? 
```{r}
tibble("all the years" = head(fvar_temp_rcp85_A), 
       "recreate" = head(fvar_temp_rcp85)) %>%  
    mutate(diff = `all the years` - `recreate`) %>% 
    knitr::kable()
```

Hmm the amount of variance explained by the PCs are very similar. 


What do the first 3 PCs look like? Do they change? 

```{r}
plot_pcs(pca_temp_rcp85_A, 1:3, title = 'Hector Emission Driven Tgav PCA RCP 8.5 Only') + 
    labs(subtitle = 'All the Years') -> 
    plot_pc_rcp85_A

plot_pc_rcp85 + 
    geom_line(data = plot_pc_rcp85_A$data, 
              aes(year, value, color = 'All the Years', linetype = pc))

```

**Conclusion:** Interesting the blue lines from the recreate RPL PCA are flatter than the PCs from all the years. The general shape is retained, just not as extreeme. Could this be because of the there is more related data going into the PCA? 

<br>

## B. Increase the size of the Hector results ensemble. 

**Expectation:** The results should not change that much

```{r}
pca_temp_rcp85_B <- prcomp(temp_rcp85_results, center = TRUE, scale. = FALSE)

# Plot the variance, I think that the majority of the variance shoudl still be captured in the first 3 PCs. 
fvar_temp_rcp85_B <- calc_variance(pca_temp_rcp85_B)
head(fvar_temp_rcp85_B)

plot_frac(fvar_temp_rcp85_B, title = 'Hector Emission Driven PCA RCP 8.5 Only All Years and All Results') + 
    coord_cartesian(xlim = c(0, 15))

```

How do the PCs change? 

```{r}
plot_pcs(pca_temp_rcp85_B, pc_indx = 1:3, title = 'Hector Emission Drive RCP 8.5') + 
    coord_cartesian(ylim = c(-.3, .4)) + 
    labs(subtitle = 'All the Years and Full Ensmble')  -> 
    more_runs

plot_pc_rcp85_A +
    geom_line(data = more_runs$data, 
              aes(year, value, color = 'increase ensemble size', linetype = pc)) + 
    labs(title = 'Compare All the Years with the Full Ensemble')
    
```


Which PC changes the most when the Hector ensemble size increases? 

```{r}
plot_pc_rcp85_A$data %>% 
    select(pc, value, year) %>% 
    mutate(id = 'small_ensemble') %>% 
    bind_rows(more_runs$data %>% 
                  select(pc, value, year) %>% 
                  mutate(id = 'full_ensemble')) %>% 
    spread(id, value) %>%  
    mutate(diff = small_ensemble - full_ensemble) %>%  
    ggplot() + 
    geom_line(aes(year, diff, color = pc)) + 
    labs(title = 'Small Ensemble - Full Ensemble PCs')


```


**Conclsion:** There are small changes in the PCs, I think that they are nearly identical to one another but I don't know what is too different from one another to matter... 

<br>

## C. 2 RCP Scenarios 

What happens when we include a second scenario?? 

So lets combine temp for RCP 85 and 26

**Expectations:** That there more PCs will contribute to the variance and that the PCs will retain similar shapes may be?


```{r, message = FALSE}

# Start by importing, checking and formatting the emissions driven run results. 
paths <- list.files(file.path(BASE, 'data-raw', 'PCA_hector_ensemble'), 'emission_rcp85|emission_rcp26', 
                    full.names = TRUE)

wide_data <- foreach(input = paths, .combine = 'full_join') %do% {
    
    # Import the data 
    str <- readRDS(input)
    
    # Format the scenario name a bit
    scn_name <- gsub(x = gsub(x = basename(input), pattern = '.rds', replacement = ''), pattern = '_', replacement = '~') 
    
    # Foramt into as a wide tibble where one row represents on "case" of Hector.
    tibble::as_tibble(str$rslt) %>% 
        filter(variable == 1) %>% 
        mutate(index = paste0(scn_name, '_', variable, '_', year)) %>% 
        select(run_id, index, value) %>% 
        spread(index, value) 
    
}

# Remove the run_id column so that the tibble only contains data that will be used in the PCA
temp_rcp85_rcp26 <- na.omit(select(wide_data, -run_id))

# Run the PCA on the data frame that contains the 2 scenarios
pca_temp_rcp85_rcp26 <- prcomp(x = temp_rcp85_rcp26, center = TRUE, scale = FALSE)
```

The fraction of the variance. 

```{r}
frac_C <- calc_variance(pca_temp_rcp85_rcp26)

plot_frac(frac_C, title = 'Hector Temp RCP 85 & Temp 85') + 
    coord_cartesian(xlim = c(0, 15)) + 
    labs(substitle = 'I only ploted the first 15 PCs')

```

Interesting I would have expected that more PCs contribute to the variance. How does it compare to the other fraction of variance?

```{r}
# The fraction of the variance for the temp rcp 85 & rcp 85 pca 
tibble::tibble( PC = 1:10,
                `temp 85 & 26` = frac_C[1:10], 
                `temp 85 only` = fvar_temp_rcp85_B[1:10]) %>%  
    knitr::kable()

```


Plot the first 5 PCs of the rcp 85 and rcp 26, I included at horizontal line at 2006 which where Hector swtiches from the historical to the rcp emissions. 

```{r}
unscaled_pca_rcp85_rcp26 <- plot_pcs(pca_temp_rcp85_rcp26, pc_indx = 1:5) 

unscaled_pca_rcp85_rcp26 + 
    geom_vline(aes(xintercept = 2006)) + 
    facet_wrap('pc') 
```

### What happens when we use prcomp to scale the PCA? 

```{r}
pca_temp_rcp85_rcp26_scaled <- prcomp(x = temp_rcp85_rcp26, center = TRUE, scale = TRUE)

plot_pcs(pca_temp_rcp85_rcp26_scaled, pc_indx = 1:5) + 
    geom_vline(aes(xintercept = 2006)) + 
    facet_wrap('pc') ->
    sclaed_pca

unscaled_pca_rcp85_rcp26$data %>% 
    mutate(scaled = FALSE) -> 
    unscaled_data 

unscaled_pca_rcp85_rcp26 %+% unscaled_data + 
    geom_line(data = sclaed_pca$data %>% 
                  mutate(scaled = TRUE),
              aes(year, value,  color = scenario, linetype = pc), 
              size = 1) + 
    facet_grid(`pc` ~ `scaled`) + 
    labs(title = 'Comparison of the unscaled (FALSE) and the scaled (TRUE)\nPCA for Hector Temp RCP 85 and RCP 26')
```

**Conclusions:** As we expected when we increase the numer of scenarios going into the PCA less variance is explained by the first 3 PCs. The sahpes are somewhat similar to one another and diverge when the rcp emissions take over. Using the prncomp scale argument changed the shape a bit, I wasn't expecting that... Where you? 

<br>

## D. RCP 85 only but now let's do CO2 and Tgav

**Expectation:**  More PCs are going to contribute to the variance, I think that the PC for the CO2 and Tgav might be smiliar but on different scales.

```{r}

# Start by importing, checking and formatting the emissions driven run results. 
paths <- list.files(file.path(BASE, 'data-raw', 'PCA_hector_ensemble'), 'emission_rcp85', full.names = TRUE)

wide_data <- foreach(input = paths, .combine = 'full_join') %do% {
    
    # Import the data 
    str <- readRDS(input)
    
    # Format the scenario name a bit
    scn_name <- gsub(x = gsub(x = basename(input), pattern = '.rds', replacement = ''), pattern = '_', replacement = '~') 
    
    # Foramt into as a wide tibble where one row represents on "case" of Hector.
    tibble::as_tibble(str$rslt) %>% 
        mutate(index = paste0(scn_name, '_', variable, '_', year)) %>% 
        na.omit %>% 
        select(run_id, index, value) %>% 
        spread(index, value) 
    
}

# Remove the run_id column so that the tibble only contains data that will be used in the PCA
temp_co2_rcp85 <- na.omit(select(wide_data, -run_id))

# Run the PCA on the data frame that contains the 2 scenarios
pca_temp_co2_rcp85 <- prcomp(x = temp_co2_rcp85, center = TRUE, scale = FALSE)
```

Because of Tgav and CO2 are on different scales I se the prcomp scale argument to be equal to TRUE. 

```{r, fig.width=10, fig.height=6}
frac_temp_co2_rcp85 <- calc_variance(pca_temp_co2_rcp85)

frac_temp_co2_rcp85[1:10]


plot_frac(frac_temp_co2_rcp85) + 
    coord_cartesian(xlim = c(0, 25)) + 
    labs(title = 'Hector RCP 85 Temp and CO2', 
         subtitle = 'unscaled & zoomed in on first 25 pcs')

plot_pcs(pca = pca_temp_co2_rcp85, pc_indx = 1:3) + 
    facet_wrap('variable', scales = 'free') + 
    labs(title = 'Hector RCP 8.5 CO2 and Tgav PCs')

``` 

**Conclusion:** I think that the story line still applies here. The first PC tells us something about warming at the begning vs the end. It makes sense to me that the wieght of the PC is greater for the CO2 vs the Tgav since there is an expontial increase in CO2 where are more of a linear incrase in Tgav. But the 2 other PCs kind of changed in shape. 


# Summary 

1. Increasing the ensmble size does not really change the PCs or the variance explained by the PCs. 
2. Incorperating more of the years flattens out the PC -- why? 
3. I think that the first PC story line is consistent across the different scenarios and variable but I am not sure about the other PCs 


## Questions 

* Why do we want to have multiple scenarios contributing to a Hector vector PCA? When we are doing the MCMC won't the likelihood function be comparing the PC from a single scenario at at time? I mean we aren't ever doing to be calibrating to muliple rcps at a time will we? 
* What do we want to do about the scaling? Do we want to remove the variance before the PCA? Because there is no way for prcomp to know the difference in the groups... the same about of variance is going to be removed from everything.
* Why do we muliply by the sign? 
