---
title: "Experiment 1"
output: html_notebook
---

## Setup 

This experiment will require the results of the monte-carlo runs produced by the
`experiment-1*.R` programs in this directory.  Those programs produce the
`*.rds` files loaded in this section.

```{r loaddata}
library('hectorcal')
library('hector')
library('coda')
library('dplyr')
library('foreach')

mcrslts_mesa_full <- readRDS('mcrslts_mesa_full.rds')
mcmc_mesa_full <- metrosamp2coda(mcrslts_mesa_full)

mcrslts_hi10 <- readRDS('mcrslts_hi10.rds')
mcmc_hi10 <- metrosamp2coda(mcrslts_hi10)

ncore <- 8
doParallel::registerDoParallel(cores=ncore)

```

## Introduction

In the runs that we did for the poster back in the fall, it looked as if the
runs might not be covering the full range of the CMIP results.  Moreover, there
seemed to be a marked falloff in the density closer to the edge of the CMIP
range, despite the fact that the likelihood function was flat over much of this
range.  A third observation was that the hector outputs for most of the sampled
parameters seemed to be approximately parallel to one another; there were very
few runs, for example, that started high and finished low, or vice versa.

All of these observations were made using the spaghetti plots of hector outputs
for a subsample of the Monte Carlo parameter samples, which is not a very precise
way of diagnosing these effects.  Therefore, the purpose of this experiment is to 
examine these matters more carefully to ensure that the Monte Carlo sampling is 
behaving as expected.

## Part A: Do the results cover the full range?

In this part we reran the full-range Monte Carlo calculations.  The code for 
performing this run is in `experiment-1A.R`; the results were saved as 
`mcrslts_mesa_full.rds`.

We were having trouble getting good mixing in these calculations.  Different 
chains were converging to different, non-overlapping regions of the parameter
space.  We were able to mitigate this problem by increasing the scale factor 
for the proposal distribution.  This had the side effect of introducing a _lot_
of autocorrelation into the Markov chains, necessitating longer runs than we
had anticipated.  Even with 8 chains of 10,000 samples each, our effective 
sample size is only a few hundred.  Together with the Gelman-Rubin diagnostic,
this suggests that our production runs could stand to be a bit longer; however,
these results should suffice for the purposes of this experiment.

```{r diag.1a}
autocorr.diag(mcmc_mesa_full)
effectiveSize(mcmc_mesa_full)
gelman.diag(mcmc_mesa_full)
```

The basic results of the calculation are below.  The marginal densities are
consistent with what we saw in the poster results.  

```{r rslt.1a}
summary(mcmc_mesa_full)
densplot(mcmc_mesa_full)
```

Do these outputs in fact cover the whole range of the CMIP outputs?  We can compare
the ranges in the comparison data to the distribution of values that come out of 
the hector runs from our sampled parameters at selected times.

```{r rangecomp}
testtimes <- c(2006, 2050, 2100)
pnames <- c(ECS(), AERO_SCALE(), DIFFUSIVITY(), BETA(), Q10_RH(), PREINDUSTRIAL_CO2())
comp_esmrcp85 <- filter(esm_comparison, experiment=='esmrcp85')
esmrcp85_ini <- system.file("input/hector_rcp85.ini", package = "hector")
hcores <- lapply(1:ncore, function(i){newcore(esmrcp85_ini)})
samps_all <- do.call(rbind, lapply(mcrslts_mesa_full, function(x){x$samples}))

filter(comp_esmrcp85, year %in% testtimes)
hector_output_stats(samps_all, hcores, pnames, nsamp=1000, 
                    times=testtimes, quantiles=c(0,0.1, 0.5, 0.9, 1))
```

The `mina` and `maxb` columns give the range of the ensemble, and as can be seen
in the quantile stats, our minimum and maximum values are actually a tiny bit
outside the CMIP range. This is because our likelihood function is not a sharp
cutoff, so the values can be a little outside the range at some times,
especially if they were in range over the rest of the run.  So, we _do_ seem to
be producing the full temperature and CO2 range.

On the other hand, looking at the 10-90 percentile range for temperature shows
that for our MC samples it is a little narrower than the corresponding range for
the CMIP models.  This indicates that the probability density is falling off as
you get to the upper and lower edge of the range, relative to what we are seeing
in the CMIP models.  We also observed this in the poster results, and it’s still
not entirely clear why that should be, since the likelihood function is mostly
flat within the CMIP range.  We do have some hypotheses; the leading one is that
the volume of the parameter space that can produce runs in these ranges is rather
small, making it hard to land in that part of the output space.

Interestingly, our 10-90 percentile range for CO2 seems to match the
corresponding CMIP range more closely than is the case for temperature.  This
observation supports the restricted volume hypothesis.  CO2 outputs depend only
on the parameters that affect the carbon cycle, while temperature outputs depend
on all six parameters jointly.  Thus, in order to get temperatures toward, say, 
the high end of the range, you need a confluence of high CO2 values and high
temperature parameters.  By contrast, to get high CO2 values, you need only 
have the right carbon cycle parameters; the temperature parameters can be anything.
The part of the parameter space that produces high temperature values is therefore
approximately a subset of the part that produces high CO2 values.

Looking at a sample of the hector output traces is also instructive.
```{r spaghetti.1a}
spaghetti_plot(mcrslts_mesa_full, 512, hcores, pnames, alpha=0.1) + ggplot2::ggtitle('Full range')
```

First, we can see that the lack of any historical constraint is allowing some
slightly absurd models into the sample.  When we rerun with our new dataset in
place, we can expect the parameter distributions to be more constrained (though
the difference may or may not be apparent on casual inspection.)  Another thing
that seems apparent is that the the concern about the lack of crossing traces, 
at least for temperature, seems to have been unfounded.  The CO2 results, on 
the other hand, show a lot less diversity in the outputs.  Evidently, the model
doesn't have as much flexibility in producing CO2 output curves with dramatically
different shapes.

## Part B: Why does the density of the traces fall off near the edge of the range?

To investigate the falloff toward the edge of the range, we set up a Monte Carlo 
run where we force the temperature to end the century in the upper decile of the 
range, while the final CO2 concentration is constrained to the full range.  No other
constraints were applied.  As might be expected, these constraints leave us with 
a more restricted set of viable parameters.

```{r compare.1a.1b}
summary(mcmc_mesa_full)   # Results from the previous run
summary(mcmc_hi10)        # Results from this run
```

As might be expected, the major difference here is that the equilibrium climate
sensitivity, $S$, is substantially higher.  There _doesn't_ seem to be any evidence
that the parameters need to be fine-tuned to land in this range of outputs.  In
particular, the interquartile range of the carbon cycle parameters isn't much 
different in the Part B run, as compared to Part A.  Likewise, the width of the 
distributions of the temperature parameters isn't much smaller than what we saw
in Part A.

Looking at the output traces, we see that constraining the output only at the 
end of the run allows some unrealistic runs into the mix.  We can also see this
in the quantiles for the aerosol scale factor, which has a much larger allowed
range on the low end.

```{r spaghetti.1b}
spaghetti_plot(mcrslts_hi10, 512, hcores, pnames, alpha=0.1) + ggplot2::ggtitle('Upper decile, final year only')
```

This difference in the early-century constraint makes it hard to definitively 
rule out the fine-tuning hypothesis.  We need to rerun this experiment with 
some full-range constraints present in the early century before we can make a
final determination on this hypothesis.
